A Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-16 03-semantics.tex
Reformatação ainda do início da Seção 3.2
Reescrevi as regras para utilizem o pacote busproofs.  Meu objetivo foi
comprimir as definições e torná-las uniformes.

Adicionei duas subsubseções novas para separar a parte que define transições
out da parte que define transições nst.

Redefini as macros das transições, agora \out e \nst em header.tex.  As
macros originais, \OUT e \NST, usam \xrightarrow, o que acaba introduzindo
um gap entre linhas adjacentes em modo texto.

% \begin{lstlisting}[
%   %numbers=left,
%   basicstyle=\ttfamily\footnotesize,
%   float=h,
%   caption={Reduced syntax of \CEU.},
%   label={lst.formal.syntax},
%   mathescape=true
% ]
%                                    // primary expressions
%   p ::= mem(id)                    (any memory access to `id')
%       $|$ awaitExt(id)               (await external event `id')
%       $|$ awaitInt(id)               (await internal event `id')
%       $|$ emitInt(id)                (emit internal event `id')
%       $|$ break                      (loop escape)
%                                    // compound expressions
%       $|$ if mem(id) then p else p   (conditional)
%       $|$ p ; p                      (sequence)
%       $|$ loop p                     (repetition)
%       $|$ every id p                 (event iteration)
%       $|$ p and p                    (par/and)
%       $|$ p or p                     (par/or)
%       $|$ fin p                      (finalization)
%                                    // derived by semantic rules
%       $|$ p @loop p                  (unwinded loop)
%       $|$ p @and q                   (unwinded par/and)
%       $|$ p @or q                    (unwinded par/or)
%       $|$ @canrun(n)                 (can run on stack level `n')
%       $|$ @nop                       (terminated expression)
% \end{lstlisting}

@@

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from an input event, how the program behaves and 
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p, n, e>\to\<p',n',e'>
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
nonnegative integers representing the current stack level,
and~$e,e'\in\E\cup\{\nil\}$ are the events emitted before and after
the transition, both possibly the empty event~($\nil$).

% \begin{align*}
% p, p' &\in\P
%     && (program~as~described~in~Listing~\ref{lst.formal.syntax})
% \\
% n, n' &\in\N
%     && (current~stack~level)
% \\
% e, e' &\in\E \cup \{\epsilon\}
%     && (emitted~event,~possibly~none)
% \end{align*}
%
At the beginning of a reaction to an input event~$id$, the triple is
initialized at stack level $0$ ($n=0$) and with the emitted event
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the triple will block with a (possibly) modified program~$p'$, at stack
level~$0$, and with no event emitted~$(\nil)$:
%
\[
  \<p,0,e>\xrightarrow{*}\<p',0,\nil>
\]

We distinguish between two types of transition rules:
    \emph{outermost} transitions $\OUT$ and
    \emph{nested} transitions $\NST$.

The $\OUT$ rules \textbf{push} and \textbf{pop} are non-recursive definitions
that only apply to the program as a whole and manipulate the stack level:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS e \neq \epsilon }
%   -----------------------------------------------------------
    { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
    & \textbf{(push)}   \\
%%%
& \frac
    { \DS n>0, \2 ((p=@nop) \vee isblocked(n,p)) }
%   -----------------------------------------------------------
    { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
    & \textbf{(pop)}    \\
%%%
%& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
    %& \textbf{(end)}    %\\
\end{eqnarray*}
}
%
Rule \textbf{push} matches whenever there is an emitted event in the triple,
and instantly broadcasts the event to the program, which means
    (a) awaking active \code{await} expressions altogether (see $bcast(p)$ in
        Figure~\ref{fig.bcast}),
    (b) creating a nested reaction by increasing the stack level, and, at the same time,
    (c) consuming the event ($e$ becomes $\epsilon$).
%
Note that rule \textbf{push} is the only one in the semantics that matches an
emitted event and also immediately consumes it.
%
The rule \textbf{pop} only decreases the stack level, not affecting the
program, and only applies if the program is blocked (see $isblocked(n,p)$ in
Figure~\ref{fig.isblocked}).
This condition ensures that an \code{emit} only resumes after its internal
reaction completes and blocks, as discussed in Section~\ref{sec.ceu.evts}.

Note that at the beginning of a reaction, an event $ext$ is emitted, which
will trigger rule \textbf{push}, which will immediately raise the stack level
to $1$.
Since rule \textbf{pop} is the only to decrease the stack level and only
applies to a blocked or terminated program, the semantics guarantees that a
reaction only terminates with a blocked program at stack level $0$.

The $\NST$ rules are recursive definitions but do not affect the stack level
and never have an emitted event as a precondition:
%
\begin{align*}
\LL p, n,\epsilon \RR &\NST
\LL p',n,e        \RR
    & \textbf{(rule-inner)}
\end{align*}
%
The distinction between $\OUT$ and $\NST$ prevents rules \textbf{push} and
\textbf{pop} to match and inadvertently modify the current stack level before
completing a nested reaction.
%
A complete reaction behaves as follows:
%
\begin{align*}
a) &\5\5
    \LL p,0,ext \RR
        \1\xrightarrow[out]{push}\1
    \LL q,1,\epsilon \RR
\\
b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
    \LL r,i,e \RR
        \1\xrightarrow[out]\1
    \LL s,j,\epsilon \RR \1]*
\\
c) &\5\5 \1\xrightarrow[in]{*}\1
    \LL t,k,\epsilon \RR
        \1\xrightarrow[out]{pop}\1
    \LL u,0,\epsilon \RR
\end{align*}
%
First (case $a$), a $\xrightarrow[out]{push}$ starts a nested reaction at level
$1$.
Then (case $b$), a series of alternations between $\NST$ transitions and a
single $\OUT$ transition represent, respectively, nested reactions and stack
operations.
Finally (case $c$), at some point, a $\xrightarrow[out]{pop}$ decrements the
stack level to $0$ and terminates the reaction.

The $\NST$ transition rules for primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL mem(id), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(mem)}        \\
%%%
\LL emit(id), n, \epsilon \RR &\NST
\LL @canrun(n), n, id \RR
    & \textbf{(emitInt)}    \\
%%%
\LL @canrun(n), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(canrun)}     \\
\end{align*}
}
%
A $mem$ operation becomes a $@nop$ which indicates the memory access (rule
\textbf{mem}).
An $emitInt(id)$ generates an event $id$ and transits to a $@canrun(n)$ which
can only resume at level $n$ (rule \textbf{emitInt}).
Since all $\NST$ rules can only transit with $e=\epsilon$, an $emitInt$ causes
rule \textbf{push} to execute at the outer level, creating a new level $n+1$ on
the stack.
Also, with the new stack level, the resulting $@canrun(n)$ itself cannot
transit, providing the desired stack-based semantics for internal events.

>>>>

%-
% \begin{lstlisting}[
%   %numbers=left,
%   basicstyle=\ttfamily\footnotesize,
%   float=h,
%   caption={Reduced syntax of \CEU.},
%   label={lst.formal.syntax},
%   mathescape=true
% ]
%                                    // primary expressions
%   p ::= mem(id)                    (any memory access to `id')
%       $|$ awaitExt(id)               (await external event `id')
%       $|$ awaitInt(id)               (await internal event `id')
%       $|$ emitInt(id)                (emit internal event `id')
%       $|$ break                      (loop escape)
%                                    // compound expressions
%       $|$ if mem(id) then p else p   (conditional)
%       $|$ p ; p                      (sequence)
%       $|$ loop p                     (repetition)
%       $|$ every id p                 (event iteration)
%       $|$ p and p                    (par/and)
%       $|$ p or p                     (par/or)
%       $|$ fin p                      (finalization)
%                                    // derived by semantic rules
%       $|$ p @loop p                  (unwinded loop)
%       $|$ p @and q                   (unwinded par/and)
%       $|$ p @or q                    (unwinded par/or)
%       $|$ @canrun(n)                 (can run on stack level `n')
%       $|$ @nop                       (terminated expression)
% \end{lstlisting}
%-

@@

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from an input event, how the program behaves and 
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
nonnegative integers representing the current stack level,
and~$e,e'\in\E\cup\{\nil\}$ are the events emitted before and after the
transition (both possibly the empty event~$\nil$).

We will refer to the triples on the left-hand and right-hand sides of
symbol~$\to$ as \emph{descriptions} (denoted~$\delta$).  The triple on the
left-hand side of symbol~$\to$ is called the \emph{input description}, and
the triple on its right-hand side is called the \emph{output description}.

%-
% \begin{align*}
% p, p' &\in\P
%     && (program~as~described~in~Listing~\ref{lst.formal.syntax})
% \\
% n, n' &\in\N
%     && (current~stack~level)
% \\
% e, e' &\in\E \cup \{\epsilon\}
%     && (emitted~event,~possibly~none)
% \end{align*}
%-

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the emitted event
($e=id$).
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a (possibly) modified program~$p'$, at stack
level~0, and with no event emitted~($\nil$):
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

We distinguish between two types of transition rules:
    \emph{outermost} transitions $\out$ and
    \emph{nested} transitions $\nst$\,.

\subsubsection*{Outermost transitions}

The~$\out$ rules \R{push} and \R{pop} are non-recursive definitions
that only apply to the program as a whole and manipulate the stack level:
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[2\jot]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$\ceu{p=\Nop}\vee\isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

%-
% { \setlength{\jot}{20pt}
% \begin{eqnarray*}
% & \frac
%     { \DS e \neq \epsilon }
% %   -----------------------------------------------------------
%     { \DS \LL p,n,e \RR \OUT \LL bcast(p),n+1,\epsilon \RR }
%     & \textbf{(push)}   \\
% %%%
% & \frac
%     { \DS n>0, \2 ((p=@nop) \vee isblocked(n,p)) }
% %   -----------------------------------------------------------
%     { \DS \LL p,n,\epsilon \RR \OUT \LL p,n-1,\epsilon \RR }
%     & \textbf{(pop)}    \\
% %%%
% %& \LL p,0,\epsilon \RR \1\xrightarrow\1 \bot
%     %& \textbf{(end)}    %\\
% \end{eqnarray*}
% }
%-

Rule \R{push} matches whenever there is an emitted event in the input
description,
and instantly broadcasts the event to the program, which means
    (a)~awaking active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ expressions altogether (see function~$\bcast$ in
        Figure~\ref{fig.bcast}),
    (b)~creating a nested reaction by increasing the stack level, and, at the same time, and
    (c)~consuming the event ($e$ becomes~$\nil$).
%
Rule \R{push} is the only rule in the semantics that matches an
emitted event and also immediately consumes it.

Rule \R{pop} only decreases the stack level, not affecting the
program, and only applies if the program is blocked (see function~$\isblocked$ in
Figure~\ref{fig.isblocked}).
This condition ensures that an $\ceu{\EmitInt}$ only resumes after its internal
reaction completes and blocks, as discussed in Section~\ref{sec.ceu.evts}.

At the beginning of the reaction, an external event is emitted, which
will trigger rule \textbf{push}, which will immediately raise the stack level
to~1.
At the end of the reaction, the program will block or terminate and
successive applications of
rule~\R{pop} will eventually lead to a description containing this
same program at stack level~0.
(Rule \R{pop} is the only rule that decreases the stack level.)

\subsubsection*{Nested transitions}

The~$\nst$ rules are recursive definitions with the following general
format:
\[
\<p,n,\nil>\nst\<p',n,e>.
\]
%
%-
% \begin{align*}
% \LL p, n,\epsilon \RR &\NST
% \LL p',n,e        \RR
%     & \textbf{(rule-inner)}
% \end{align*}
%-
%
Nested transitions do not affect the stack level and never have an emitted
event as a precondition.  The distinction between~$\out$ and~$\nst$ prevents
rules \R{push} and \R{pop} from matching and, consequently, from
inadvertently modifying the current stack level before the nested reaction
is complete.

A complete reaction consists of the following series of transitions:
\begin{align*}
  \<p,0,e_\ext>\out[push]\<p_1,1,\nil>
  \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
  \null\nst[*]\null\out[pop]\<p',0,\nil>\,.
\end{align*}
%
%-
% \begin{align*}
% a) &\5\5
%     \LL p,0,ext \RR
%         \1\xrightarrow[out]{push}\1
%     \LL q,1,\epsilon \RR
% \\
% b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
%     \LL r,i,e \RR
%         \1\xrightarrow[out]\1
%     \LL s,j,\epsilon \RR \1]*
% \\
% c) &\5\5 \1\xrightarrow[in]{*}\1
%     \LL t,k,\epsilon \RR
%         \1\xrightarrow[out]{pop}\1
%     \LL u,0,\epsilon \RR
% \end{align*}
%-
%
First, a~$\out[push]$ starts a nested reaction at level~1.
Then, a series of alternations between zero or more~$\nst$ transitions (nested reactions) and a
single~$\out$ transition (stack operation) takes place.
Finally, a last~$\out[pop]$ transition decrements the
stack level to~0 and terminates the reaction.

The~$\nst$ transition rules for atomic expressions are defined as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL mem(id), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(mem)}        \\
%%%
\LL emit(id), n, \epsilon \RR &\NST
\LL @canrun(n), n, id \RR
    & \textbf{(emitInt)}    \\
%%%
\LL @canrun(n), n, \epsilon \RR &\NST
\LL @nop, n, \epsilon \RR
    & \textbf{(canrun)}     \\
\end{align*}
}
%
A $mem$ operation becomes a $@nop$ which indicates the memory access (rule
\textbf{mem}).
An $emitInt(id)$ generates an event $id$ and transits to a $@canrun(n)$ which
can only resume at level $n$ (rule \textbf{emitInt}).
Since all $\NST$ rules can only transit with $e=\epsilon$, an $emitInt$ causes
rule \textbf{push} to execute at the outer level, creating a new level $n+1$ on
the stack.
Also, with the new stack level, the resulting $@canrun(n)$ itself cannot
transit, providing the desired stack-based semantics for internal events.
